Follow-up Samples

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
add discussion changes

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  We chose the middle training model, based on AUC, from 100 80/20 (train/test) splits.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using this model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample as a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.  Although there was a decrease in overall probability of an adenoma or carcinoma only `r followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Only a single OTU was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  This research provides evidence that it is possible to create a highily sensitive model for detection of adenoma or carcinoma with bacterial microbiome data.  It accomplishes this by using a unique sample set in which before and after stool samples are available for assessment.  By using these types of samples we are able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool (**insert citation**).  Based on previous work by Jobin, et al. (**insert citation**)  it may not be surprising to see E.coli in the top 5 OTUs for the training models.  Similarily, Porphyromonas has also been implicated in colorectal cancer (**insert citation**).  Interestingly, the other OTUs had taxonomic identification for resident gut microbes.  This could suggest that changes to the resident microbiome are important to the initiation of adenoma or carcinoma formation (**insert citation**) and provide support for the hypothesis that an initial change in the bacterial microbiome could pave the way for more inflammatory species: whether by creation of a new niche for oral microbes (**insert citation**) or allowing for a bloom of existing pro-inflammatory residents (**insert citation**).

Naturally, it is curious that normal staples of many screening studies such as Fusobacterium, Parvimonas, and Peptostreptococcus were not present in the majority of the training models.  One potential explanation for this is that FIT provides the same information to the model as these three organisms and so the model uses FIT preferentially over them.  This has been suggested to be the case in a previous study (**insert Baxter Study**).  Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the best model does not utilize these specific bacteria for prediction purposes.  Another potential reason why we did not identify the usual suspects is that they may not change much between our initial and follow up samples.  That is to say that they are consistently present even after removal of the lesion but surgery.

One limitation is that we do not know whether these individuals eventually had a CRC remission.  This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having CRC or adenomas.  Another limitation is that we do not know if adding modern tests such as the stool DNA test (**insert citation**) could help improve our overall AUC.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing that it can successfully predict both carcinoma and adenoma our study provides further data that these patient factors in conjunction with the bacterial microbiome could potentially influence CRC and perhaps have a role in formation of adenomas.  Further studies need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations.

Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome.  The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures associated with these lesions.  It also shows that these algorithms successfully react to successful treatment regimens and may be able to one day diagnose CRC with a high level of accuracy.

>>>>

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  Our middle training model, based on AUC, from 100 80/20 (train/test) splits was similar to the full training data model.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using the full training data model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample was a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.  Although there was a decrease in overall probability of an adenoma or carcinoma only `r followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Many of the OTUs identified classified to normal flora bacterium [Table S4].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool (**insert citation**).  Based on previous work by Jobin, et al. (**insert citation**)  it may not be surprising to see E.coli in the top 5 OTUs for this model.  Similarily, Porphyromonas has also been implicated in colorectal cancer (**insert citation**).  Interestingly, many of the other OTUs had taxonomic identification for resident gut microbes.  This could suggest that changes to the resident microbiome are important to the initiation of adenoma or carcinoma formation (**insert citation**) and provide support for the hypothesis that an initial change in the bacterial microbiome could pave the way for more inflammatory species: whether by creation of a new niche for oral microbes (**insert citation**) or allowing for a bloom of existing pro-inflammatory residents (**insert citation**).

Naturally, it is curious that normal staples of many screening studies such as Fusobacterium, Parvimonas, and Peptostreptococcus were not present in the majority of the training models.  One potential explanation for this is that FIT provides the same information to the model as these three organisms and so the model uses FIT preferentially over them.  This has been suggested to be the case in a previous study (**insert Baxter Study**).  Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the model does not utilize these specific bacteria for prediction purposes.  Another potential reason why we did not identify the "usual suspects"" is that these bacteria may not change much between initial and follow up samples in those with an identified lesion.  That is to say that the bacteria are consistently present even after removal of the lesion by surgery.  Finally, it is likely that within our test set there was not enough indviduals in which detection  was made or relative abundance high enough for these bacteria to be significant using a paired wilcoxson test.

One limitation in this study is that we do not know whether individuals in our test set eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test (**insert citation**) could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing that it can successfully help to predict both carcinoma and adenoma our study provides further data that these patient factors in conjunction with the bacterial microbiome could potentially influence CRC and perhaps have a role in formation of adenomas.  Further studies need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome.  The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures associated with these lesions.  It also shows that these algorithms can not only successfully react to successful treatment regimens but also may be able to one day diagnose CRC with a high level of accuracy.

### 2016-12-07 submission/manuscript_outline_20161024.tex
add discussion changes

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). We chose the middle training model, based on AUC, from 100
80/20 (train/test) splits. It's 10-fold cross validated AUC was similar
to it's test set AUC which was not the case for both the best and worse
training model {[}Figure 3{]}. Using this model we predicted the
probability of a lesion in the inital and follow up samples {[}Figure
4{]}. There was a signficant decrease in positive probability regardless
of whether the sample as a carcinoma or adenoma. The overall sensitivity
for lesion detection in the intial samples was 100 and for follow ups
was 100. Although there was a decrease in overall probability of an
adenoma or carcinoma only 24 were below the 0.5 threshold out of the
total 65 individuals who were diagnosed as not having a carcinoma on
follow up.

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S4{]}. Only a single OTU was significant
after multiple comparison correction and the lowest taxonomic
identification of Otu000012 was to Blautia. Although there was a
difference in the relative abundance at initial and follow up these
values were not drastically different from the relative abundance values
observed in the control individuals of the training set {[}Figure 5{]}.
This research provides evidence that it is possible to create a highily
sensitive model for detection of adenoma or carcinoma with bacterial
microbiome data. It accomplishes this by using a unique sample set in
which before and after stool samples are available for assessment. By
using these types of samples we are able to show that this model is
reactive. That is to say that after surgery for removal of the adenoma
or carcinoma it decreases the positive probability to reflect a lower
likelihood of the individual having an adenoma or carcinoma.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
(\textbf{insert citation}). Based on previous work by Jobin, et al.
(\textbf{insert citation}) it may not be surprising to see E.coli in the
top 5 OTUs for the training models. Similarily, Porphyromonas has also
been implicated in colorectal cancer (\textbf{insert citation}).
Interestingly, the other OTUs had taxonomic identification for resident
gut microbes. This could suggest that changes to the resident microbiome
are important to the initiation of adenoma or carcinoma formation
(\textbf{insert citation}) and provide support for the hypothesis that
an initial change in the bacterial microbiome could pave the way for
more inflammatory species: whether by creation of a new niche for oral
microbes (\textbf{insert citation}) or allowing for a bloom of existing
pro-inflammatory residents (\textbf{insert citation}).

Naturally, it is curious that normal staples of many screening studies
such as Fusobacterium, Parvimonas, and Peptostreptococcus were not
present in the majority of the training models. One potential
explanation for this is that FIT provides the same information to the
model as these three organisms and so the model uses FIT preferentially
over them. This has been suggested to be the case in a previous study
(\textbf{insert Baxter Study}). Regardless, our study does not argue
against the importance of these bacterium in CRC initiation or
pathogenesis but rather that the best model does not utilize these
specific bacteria for prediction purposes. Another potential reason why
we did not identify the usual suspects is that they may not change much
between our initial and follow up samples. That is to say that they are
consistently present even after removal of the lesion but surgery.

One limitation is that we do not know whether these individuals
eventually had a CRC remission. This information would help to
strengthen the case for our Random Forest based model keeping a number
of individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having CRC or adenomas. Another limitation
is that we do not know if adding modern tests such as the stool DNA test
(\textbf{insert citation}) could help improve our overall AUC. Finally,
although our training and test set are relatively large we still run the
risk of overfitting or having a model that may not be immediately
extrapolateable to other populations. We've done our best to safeguard
against this by not only running 10-fold cross validation but also
having over 100 different 80/20 splits to try and mimic the type of
variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing
that it can successfully predict both carcinoma and adenoma our study
provides further data that these patient factors in conjunction with the
bacterial microbiome could potentially influence CRC and perhaps have a
role in formation of adenomas. Further studies need to be carried out to
verify our findings since not only are we dealing with stool, which
could be very different than the communities present on the actual
tissue, but also are dealing with correlations.

Despite these limitations we think that these findings significantly add
to the existing scientific knowledge on CRC and the bacterial
microbiome. The ability for machine learning algorithms to take
bacterial microbiome data and successfully lower positive probability
after either adenoma or carcinoma removal provides evidence that there
are specific signatures associated with these lesions. It also shows
that these algorithms successfully react to successful treatment
regimens and may be able to one day diagnose CRC with a high level of
accuracy.

>>>>

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). Our middle training model, based on AUC, from 100 80/20
(train/test) splits was similar to the full training data model. It's
10-fold cross validated AUC was similar to it's test set AUC which was
not the case for both the best and worse training model {[}Figure 3{]}.
Using the full training data model we predicted the probability of a
lesion in the inital and follow up samples {[}Figure 4{]}. There was a
signficant decrease in positive probability regardless of whether the
sample was a carcinoma or adenoma. The overall sensitivity for lesion
detection in the intial samples was 100 and for follow ups was 100.
Although there was a decrease in overall probability of an adenoma or
carcinoma only 24 were below the 0.5 threshold out of the total 65
individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S4{]}. Many of the OTUs identified
classified to normal flora bacterium {[}Table S4{]}. Only a single OTU
though was significant after multiple comparison correction and the
lowest taxonomic identification of Otu000012 was to Blautia. Although
there was a difference in the relative abundance at initial and follow
up these values were not drastically different from the relative
abundance values observed in the control individuals of the training set
{[}Figure 5{]}. This research provides evidence that it is possible to
use bacterial microbiome data to create a highily sensitive model, that
is reactive to therapy, for detection of adenoma or carcinoma. It
accomplishes this by using a unique sample set in which before and after
surgery stool samples are available for assessment. By using these types
of samples we are not only able to show sensitivity of lesion prediction
but also able to show that this model is reactive. That is to say that
after surgery for removal of the adenoma or carcinoma it decreases the
positive probability to reflect a lower likelihood of the individual
having an adenoma or carcinoma.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
(\textbf{insert citation}). Based on previous work by Jobin, et al.
(\textbf{insert citation}) it may not be surprising to see E.coli in the
top 5 OTUs for this model. Similarily, Porphyromonas has also been
implicated in colorectal cancer (\textbf{insert citation}).
Interestingly, many of the other OTUs had taxonomic identification for
resident gut microbes. This could suggest that changes to the resident
microbiome are important to the initiation of adenoma or carcinoma
formation (\textbf{insert citation}) and provide support for the
hypothesis that an initial change in the bacterial microbiome could pave
the way for more inflammatory species: whether by creation of a new
niche for oral microbes (\textbf{insert citation}) or allowing for a
bloom of existing pro-inflammatory residents (\textbf{insert citation}).

Naturally, it is curious that normal staples of many screening studies
such as Fusobacterium, Parvimonas, and Peptostreptococcus were not
present in the majority of the training models. One potential
explanation for this is that FIT provides the same information to the
model as these three organisms and so the model uses FIT preferentially
over them. This has been suggested to be the case in a previous study
(\textbf{insert Baxter Study}). Regardless, our study does not argue
against the importance of these bacterium in CRC initiation or
pathogenesis but rather that the model does not utilize these specific
bacteria for prediction purposes. Another potential reason why we did
not identify the ``usual suspects''" is that these bacteria may not
change much between initial and follow up samples in those with an
identified lesion. That is to say that the bacteria are consistently
present even after removal of the lesion by surgery. Finally, it is
likely that within our test set there was not enough indviduals in which
detection was made or relative abundance high enough for these bacteria
to be significant using a paired wilcoxson test.

One limitation in this study is that we do not know whether individuals
in our test set eventually had a subsequent CRC diagnosis. This
information would help to strengthen the case for our Random Forest
based model keeping a number of individuals above the cutoff threshold
even though at follow up they were diagnosed as no longer having a
lesion. Another limitation is that we do not know if adding modern tests
such as the stool DNA test (\textbf{insert citation}) could help improve
our overall AUC. Another limitation is that this study drew heavily from
those with caucasian ancestry. The results may not be immediately
representative of those with either Asian or African ancestry. Finally,
although our training and test set are relatively large we still run the
risk of overfitting or having a model that may not be immediately
extrapolateable to other populations. We've done our best to safeguard
against this by not only running 10-fold cross validation but also
having over 100 different 80/20 splits to try and mimic the type of
variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing
that it can successfully help to predict both carcinoma and adenoma our
study provides further data that these patient factors in conjunction
with the bacterial microbiome could potentially influence CRC and
perhaps have a role in formation of adenomas. Further studies need to be
carried out to verify our findings since not only are we dealing with
stool, which could be very different than the communities present on the
actual tissue, but also are dealing with correlations that may not be
representative of the true pathogensis of disease.

Despite these limitations we think that these findings significantly add
to the existing scientific knowledge on CRC and the bacterial
microbiome. The ability for machine learning algorithms to take
bacterial microbiome data and successfully lower positive probability
after either adenoma or carcinoma removal provides evidence that there
are specific signatures associated with these lesions. It also shows
that these algorithms can not only successfully react to successful
treatment regimens but also may be able to one day diagnose CRC with a
high level of accuracy.
